---
title: "QBS103 Project - Elodie Richard"
output:
  word_document: default
  pdf_document: default
date: "2024-07-25"
---
```{r}
project1.data <- list.files(path = "/Users/elodierichard/Documents/QBS103/Project Submission 1 Data", pattern = ".csv")
print(project1.data) #I first moved both data files into one folder on my laptop to retrieve it
```
```{r}
setwd("/Users/elodierichard/Documents/QBS103/Project Submission 1 Data") #this is to set the working directory to the data files in this folder

genes <- read.csv("QBS103_GSE157103_genes.csv") #this is to rename and retrieve the first gene data file
the_matrix <- read.csv("QBS103_GSE157103_series_matrix.csv") #this is to rename and retrieve the second series matrix data file

head(genes) #this is used to visualize the data and only for 6 rows
head(the_matrix)
```

```{r}
#creating a genes data frame for the genes file
test_genes <- as.data.frame(t(genes)) 
names(test_genes) <- test_genes[1,] #this allows the genes table to be organized according to names by adding an extra row with the names
test_genes <- test_genes[-1,] #this removes the first row containing "x" in the genes file so that it can be combined in the next step with the matrix file
```

```{r}
test_genes$participant_id <- row.names(test_genes) #this will move the participant id into it's own column in the table
combined <- merge(test_genes, the_matrix, by = 'participant_id') #this combines the genes file and matrix file together to create one table that I names "combined"
```

```{r}
#Histogram using the gene AAAS 
library(ggplot2)
setwd("/Users/elodierichard/Documents/QBS103/Project Submission 1 Data")

combined$ABCA13 <- as.numeric(combined$ABCA13) #needed to make them all as.numeric for submission 2
combined$AACS <- as.numeric(combined$AACS)
combined$AAAS <- as.numeric(combined$AAAS) #this is so that the plot can pull just the gene AAAS from the combined data in order to plot it

histogram <- ggplot(combined, aes(x=combined$AAAS)) + #this called on ggplot to use the file "combined" then aes was used to plot the x-axis with the gene chosen AAAS
  geom_histogram(bins = 20, color = 'navy', fill = 'lightblue') +  #this generated the histogram with the number of bars (20), the color and fill of each bar
  labs(title = 'Gene Expression of AAAS', #this labeled the title and the axis
        x= 'Gene: AAAS' , 
        y= 'Frequency of AAAS' )
plot(histogram)
```

```{r}
#Scatterplot of the gene expression of AAAS compared to ferritin levels
library(ggplot2)
combined$ferritin.ng.ml. <- as.numeric(combined$ferritin.ng.ml.) #used to pull out ferritin to plot 
#comments mostly the same as for histogram except for a few changes
scatterplot <- ggplot(combined, aes(x= combined$ferritin.ng.ml., y = combined$AAAS)) + #need to specifiy what is on the y-axis
  geom_point(bins = 10, color = 'violet') +  #use geom_point for a scatter plot to be generated
  labs(title = 'Gene Expression of AAAS vs. Ferritin Levels' , 
       x= 'Ferritin Levels (ng/mL)', 
       y= 'Gene Expression of AAAS')
#for trendline it's geom_smooth
plot(scatterplot)
```

```{r}
#Scatterplot for Gene Expression vs Age (this was run to compare different data to see differences) not using for presentation
library(ggplot2)

scatterplot_practice<- ggplot(combined, aes(x= combined$age, y = combined$AAAS)) + 
  geom_point(bins = 10, color = 'green') +  
  labs(title = 'Gene Expression of AAAS vs. Age' , x= 'Age of Participant (yrs)', y= 'Gene AAAS')
plot(scatterplot_practice)
```

```{r}
#Boxplot comparing gene expression of AAAS related to ICU status depending on Age
library(ggplot2)
#similar process to histogram and scatterplot with a few adjustments
boxplot <- ggplot(combined, aes(x=icu_status, y = AAAS, fill = age)) + #need to add a fill to demonstrate the age range depending on gene expression and if ICU status
  geom_boxplot(bins = 10, fill = 'maroon') + #to generate a box plot use geom_boxplot
  labs(title = 'Gene Expression of AAAS vs ICU Status and Participant Age',  #to label each attribute of the boxplot
       x= 'ICU Status of Participant' , 
       y= 'Gene Expression of AAAS', 
       fill= 'Age of Participant (yrs)')
#plot(boxplot)
```
```{r}
### this is the fixed boxplot from the previous submission so that it includes a categorical variable (mechanical ventilation) instead of another continuous variable (age)
#Boxplot comparing gene expression of AAAS related to ICU status depending on Age
library(ggplot2)
#similar process to histogram and scatterplot with a few adjustments
boxplot <- ggplot(combined, aes(x=icu_status, y = AAAS, fill = mechanical_ventilation)) + #need to add a fill to demonstrate the age range depending on gene expression and if ICU status
  geom_boxplot() + #to generate a box plot use geom_boxplot
  scale_fill_manual(values = c('pink','salmon')) +
  labs(title = 'Gene Expression of AAAS vs ICU Status and Mechanical Ventilation',  #to label each attribute of the boxplot
       x= 'ICU Status of Participant' , 
       y= 'Gene Expression of AAAS', 
       fill= 'Mechanical Ventilation')
plot(boxplot)
```

Submission 2
Build a function to create the plots you made for Presentation 1, incorporating any feedback you received on your submission. Your functions should take the following input: (1) the name of the data frame, (2) a list of 1 or more gene names, (3) 1 continuous covariate, and (4) two categorical covariates (10 pts)
Select 2 additional genes (for a total of 3 genes) to look at and implement a loop to generate your figures using the function you created (10 pts)
Present one of your boxplots in class. Be prepared to explain the gene and covariates you chose and comment on the distribution as if you were presenting your research findings. No slides are required, just bring your plot. In class, be prepared to provide constructive feedback for your classmates (5 pts)
Make sure you push your code to your git repository prior to class. As a reminder, we do not need you to share your GitHub repository until the final submission. Pushing this submission to GitHub will be worth 5 pts on the final submission and you can earn 1 additional point on your final project grade if you push 1 extra time along the way (changes between pushes must be significant to earn the extra point). 

```{r}
## comment out things you don't need anymore

sub_2_plots <- function(data, genes, cat1 , cat2, cont ) { #created a new function with each element defined
    
    histogram2 <- ggplot(data, aes_string(x= genes)) + geom_histogram(bins = 20, color = 'navy', fill = 'lightblue') + labs(title = 'Gene Expression', x= 'Gene' , y= 'Frequency of the Gene' )
    scatterplot2 <- ggplot(data, aes_string(x= cont, y = genes)) + geom_point(bins = 10, color = 'violet') + labs(title = 'Gene Expression vs. Ferritin Levels' , x= 'Ferritin Levels (ng/mL)', y= 'Gene Expression')
   boxplot2 <- ggplot(data, aes_string(x= cat1, y = genes, fill = cat2)) + geom_boxplot() + scale_fill_manual(values = c('pink','salmon')) + labs(title = 'Gene Expression vs ICU Status and Mechanical Ventilation', x= 'ICU Status of Participant' , y= 'Gene Expression', fill= 'Mechanical Ventilation')

    plot(histogram2)
    plot(scatterplot2)
    plot(boxplot2)
}

specific_genes = subset(combined, select = c("AAAS", "AACS", "ABCA13")) #this is to subset the data table so that it only runs through the chosen genes and not every gene in the table
### had to be sure to make these chosen genes as.numeric so that it would run through
for (gene in colnames(specific_genes)) { #created a for loop to run through every gene chosen in the newly created subset of the data
  print(gene)
  print(combined[,gene]) # this will grab each row and the gene columns specifically
  sub_2_plots(combined, gene , cat1 = "icu_status", cat2 = "mechanical_ventilation", cont = "ferritin.ng.ml.")} #defined each of the variables from the function above so that it would plot

 # sub_2_plots(combined, "AAAS", cat1 = "icu_status", cat2 = "mechanical_ventilation", cont = "ferritin.ng.ml.")
```



Generate a table formatted in LaTeX of summary statistics for all the covariates you looked at and 2 additional continuous (3 total) and 1 additional categorical variable (3 total). (5 pts)
Stratifying by one of your categorical variables
Tables should report n (%) for categorical variables
Tables should report mean (sd) or median [IQR] for continuous variables
```{r}
#na.omit(combined)
sd(na.omit(combined$icu_status)[combined$icu_status == 'yes' & combined$icu_status == 'no'])
mean(na.omit(combined$icu_status)[combined$icu_status == 'yes' & combined$icu_status == 'no'])
```



```{r}
library(tidyverse)
library(knitr) # for base kable function
library(kableExtra) # for formatting 

#mean_ferritin <- 833.5
#sd_ferritin <- 1042.798
#mean_procalcitonin <- 3.08
#sd_procalcitonin <- 9.788764
#mean_lactate <- 1.46
#sd_lactate <- 1.167462



# Define Table
table1 <- data.frame('Ferritin' = c('mean (sd)','Sex n (%)',
                                    "Female","Male"),
                     "Value" = c('35 (2)','','10 (25.0)','30 (75.0)'))
table1
# Print table using kable, will only show once knitted because it's LaTeX format
tab <- kable(x = table1, caption = 'Summary Table',
             format = 'latex',booktabs = T,
             col.names = c("Variable", "n = 40"),
             align = c('l','r'),escape = T) %>%
  add_indent(positions = c(3,4),level_of_indent = 1)
tab

source('TableOne.R')

# Define data set to use
#set.seed(3927)
randomData <- data.frame(ferritin.ng.ml. = rnorm(n = 1000,mean = 833.5,sd = 1042.798),
                         procalcitonin.ng.ml = rnorm(n = 1000, mean = 3.08,sd = 9.788764),
                         lactate.mmol.l = rexp(n = 1000,rate = 0.25),
                         Cat1 = factor(rbinom(n = 1000,size = 1,prob = 0.50),
                                       labels = c(F,T)),
                         Cat2 = factor(rbinom(n = 1000,size = 1,prob = 0.25),
                                       labels = c(F,T)),
                         Cat3 = factor(rbinom(n = 1000,size = 2,prob = 0.40),
                                       labels = c('A','B','C')))


#buildTableOne(data = randomData,varList = c('Cont1','Cont2','Cat1','Cont3','Cat2','Cat3'),
            #  nonnormVars = 'Cont3',catVars = c('Cat1','Cat2','Cat3'))
```

```{r}
install.packages("tableone")
library(tableone)
```

```{r}
combined$procalcitonin.ng.ml.. <- as.numeric(combined$procalcitonin.ng.ml..) 
combined$lactate.mmol.l. <- as.numeric(combined$lactate.mmol.l.) 
project_table <- CreateTableOne(data = combined, vars = c('ferritin.ng.ml.', 'procalcitonin.ng.ml..', 'lactate.mmol.l.', 'icu_status', 'sex'), strata = 'mechanical_ventilation' )
#print(project_table)
project_table1 <- print(project_table, showAllLevels = T, nonnormal = c('ferritin.ng.ml.','procalcitonin.ng.ml..', 'lactate.mmol.l.'))
write.csv(project_table1, '/Users/elodierichard/Documents/QBS103/Project_Table1.csv')
#table1(~ferritin.ng.ml. + procalcitonin.ng.ml.. + lactate.mmol.l. + icu_status + mechanical_ventilation)
```

#Generate final a publication quality  histogram, scatter plot, and boxplot from submission 1 (i.e. only for your first gene of interest)  (5 pts)

```{r}
library(ggplot2)
setwd("/Users/elodierichard/Documents/QBS103/Project Submission 1 Data")

sub_2_plots <- function(data, genes, cat1 , cat2, cont ) { #created a new function with each element defined
    
    histogram2 <- ggplot(data, aes_string(x= genes)) + geom_histogram(bins = 20, color = 'navy', fill = 'lightblue') + labs(title = 'Gene Expression of AAAS', x= 'Gene AAAS' , y= 'Frequency of the AAAS' )
    scatterplot2 <- ggplot(data, aes_string(x= cont, y = genes)) + geom_point(bins = 10, color = 'violet') + labs(title = 'Gene Expression of AAAS vs. Ferritin Levels' , x= 'Ferritin Levels (ng/mL)', y= 'Gene Expression of AAAS')
   boxplot2 <- ggplot(data, aes_string(x= cat1, y = genes, fill = cat2)) + geom_boxplot() + scale_fill_manual(values = c('pink','salmon')) + labs(title = 'Gene Expression of AAAS vs ICU Status and Mechanical Ventilation', x= 'ICU Status of Participant' , y= 'Gene Expression of AAAS', fill= 'Mechanical Ventilation')

    plot(histogram2)
    plot(scatterplot2)
    plot(boxplot2)
}

specific_genes = subset(combined, select = c("AAAS")) #this is to subset the data table so that it only runs through the chosen genes and not every gene in the table
### had to be sure to make these chosen genes as.numeric so that it would run through
for (gene in colnames(specific_genes)) { #created a for loop to run through every gene chosen in the newly created subset of the data
  print(gene)
  print(combined[,gene]) # this will grab each row and the gene columns specifically
  sub_2_plots(combined, gene , cat1 = "icu_status", cat2 = "mechanical_ventilation", cont = "ferritin.ng.ml.")} #defined each of the variables from the function above so that it would plot

 #sub_2_plots(combined, "AAAS", cat1 = "icu_status", cat2 = "mechanical_ventilation", cont = "ferritin.ng.ml.")
```

Generate a heatmap (5 pts)
Heatmap should include at least 10 genes
Include tracking bars for the 2 categorical covariates in your boxplot
Heatmaps should include clustered rows and columns

```{r}
install.packages('pheatmap')
library(pheatmap)
```

```{r}

#head(select(participant_id, AAAS, AACS, AAGAB, AAK1, AAMDC, AAMP, AANAT, AAR2, AARS1, AARS2))

combined$AAGAB <- as.numeric(combined$AAGAB) 
combined$AAK1 <- as.numeric(combined$AAK1) 
combined$AAMDC <- as.numeric(combined$AAMDC) 
combined$AAMP <- as.numeric(combined$AAMP) 
combined$AANAT <- as.numeric(combined$AANAT) 
combined$AAR2 <- as.numeric(combined$AAR2) 
combined$AARS1 <- as.numeric(combined$AARS1) 
combined$AAAS <- as.numeric(combined$AAAS) 
combined$AACS <- as.numeric(combined$AACS)
combined$AARS2 <- as.numeric(combined$AARS2) 

head(combined)

#new_combined(combined) <- c('AAAS', 'AACS', 'AAGAB', 'AAK1', 'AAMDC', 'AAMP', 'AANAT', 'AAR2', 'AARS1', 'AARS2')

```
```{r}
# Generate heatmap without clustering
#heatmap_asnumeric <- as.numeric(combined$[2:11] margin=2, ) 

###before adding annotations

#pheatmap(combined[,c('AAAS', 'AACS', 'AAGAB', 'AAK1', 'AAMDC', 'AAMP', 'AANAT', 'AAR2', 'AARS1', 'AARS2')],
       #  show_rownames = F,
       #  cluster_rows = T,
        # cluster_cols = T)
```
```{r}
#annotationData <- data.frame(row.names = colnames(combined_df),
                            # 'Gene' = c(factor(rbinom(n = 25,size = 1,prob = 0.6),
                                      #           labels = c('ICU Status','Mechanical Ventilation'))))

combined_df <- data.frame(combined[,c('AAAS', 'AACS', 'AAGAB', 'AAK1', 'AAMDC', 'AAMP', 'AANAT', 'AAR2', 'AARS1', 'AARS2')])

annotationData <- data.frame(ICU_Status =combined$icu_status ,Mechanical_Ventilation= combined$mechanical_ventilation, row.names = row.names(combined_df))
row.names(combined_df) <- row.names(combined)
row.names(annotationData) <- row.names(combined)


annotationColors <- list(ICU_Status= c(' yes' = 'aquamarine4',
                                    ' no' = 'deepskyblue4'),
                         Mechanical_Ventilation= c(' yes' = 'cyan',
                                                   ' no' = 'blue')
                         )

combined$icu_status 
pheatmap(combined_df,
         show_rownames = F,
         cluster_rows = T,
         cluster_cols = T,
         annotation_row = annotationData,
         annotation_colors = annotationColors
         )
```

Going through the documentation for ggplot2, generate a plot type that we did not previously discuss in class that describes your data in a new and unique way (5 pts) 
```{r}
#look through ggplot to find cool new ways
install.packages('hexbin')
library(ggplot2)
#similar process to histogram and scatterplot with a few adjustments
new_plot <- ggplot(combined, aes(x=icu_status, y = AAAS, fill = mechanical_ventilation)) + #need to add a fill to demonstrate the age range depending on gene expression and if ICU status
  geom_hex() + #to generate a box plot use geom_boxplot
  scale_fill_manual(values = c('pink','salmon')) +
  labs(title = 'Gene Expression of AAAS vs ICU Status and Mechanical Ventilation',  #to label each attribute of the boxplot
       x= 'ICU Status of Participant' , 
       y= 'Gene Expression of AAAS', 
       fill= 'Mechanical Ventilation')
plot(new_plot)
```
```{r}
#look through ggplot to find cool new ways
#install.packages('ggbeeswarm')
library(ggplot2)
library(reshape2)
#similar process to histogram and scatterplot with a few adjustments
new_plot <- combined(combined[, c('icu_status','mechanical ventilation')], id='AAAS')
ggplot(new_plot) + geom_line(aes((x=status, y = AAAS, fill = variable)) + labs(title= 'Gene Expression')
#new_plot <- melt(combined, aes(x=icu_status, y = AAAS, fill = mechanical_ventilation)) + #need to add a fill to demonstrate the age range depending on gene expression and if ICU status
 # geom_b() + #to generate a box plot use geom_boxplot
 # scale_fill_manual(values = c('pink','salmon')) +
 # labs(title = 'Gene Expression of AAAS vs ICU Status and Mechanical Ventilation',  #to label each attribute of the boxplot
  #     x= 'ICU Status of Participant' , 
   #    y= 'Gene Expression of AAAS', 
     #  fill= 'Mechanical Ventilation')
plot(new_plot)
```

Submit a LaTeX file and knitted PDF file summarizing your results (20 pts total). This file should include the following sections:
Table of summary statistics
Histogram of gene
Scatter plot of gene + continuous covariate
Boxplot of gene stratified by 2 categorical covariates
Heatmap
Your selected new plot type
Introduction: Brief description of the data set and your gene of choice for your main plots. Additional gene descriptions aren’t required for genes included in your heatmap.
Methods: Brief summary of methods including data source, R version and packages, and clustering algorithm used, as discussed in class.
Results: Description of the findings of each table/figure (outlined below). While you do not need to provide an extensive analysis of each item, you must provide a brief statement referencing them and then cite the relevant table or figure, as discussed in class. Example: “Gene x did not appear to be associated with covariate y (Figure 2).” Additionally, you must typeset and provide captions/figure legends for each item as discussed in class. Required elements include:
References: At a minimum, your references must include the paper that the dataset came from, an original source for your gene description, and R packages used. All references should be cited within the text as shown in class.
Submit a link to your github repository for review (15 pts total; 5 per presentation)
Push all the clearly commented code for your final submission.
You must have a commit from before each presentation including all of the code used for each presentation.
Repository must be public facing after final submission.
Provide a brief presentation of your new plot type providing a description of what it shows and why you think it’s useful. (5 pts)
Reminder: All figures and tables should be “publication ready” (i.e. clean variable names, etc. as discussed in class). While we have not deducted points for this so far in prior iterations of this project, you will lose points on your final submission for sloppy figures and tables.
```{r}

```


